Supplementary Information: Vulnerability of tree growth

### 2025-03-03 doc/manuscript.Rmd
title options, basic changes to format
for Ecology Letters

---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
csl: new-phytologist.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

**Title:** Sensitivity of tree growth to drought in seasonally dry tropical forests using long-term dendrometer band measurements

@@

@@

# Summary {-}
<!-- 
New Phytologist Summary format
- Rationale  
- Methods  
- Key results   
- Main conclusion   -->

@@

@@

Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to climate anomalies is poorly understood.  
- Woody growth has long-lasting effects on the terrestrial carbon pool but is sensitive to interannual variation in temperature and precipitation.  
- Extreme events like drought, with temperature and precipitation outside normal ranges that can lead to growth reductions that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].  
- the extent of this variation in drought responses is poorly understood.  
- with changing climate patterns across the global tropics, including novel climatic regimes [@dahinden_future_2017], there is a need to understand the drivers of this variation, to be able to predict and respond to forest responses to drought.

@@

-High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
-Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity

@@

Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages between these processes. 
We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables. 
However, we also ran simpler models with species random effects only on the intercept, the results of which are reported in the Supplementary Material. 
Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as an monotonic predictor (ordered factor) and centred and scaled the other predictors. 
We modelled CII as an ordinal categorical variable with a cumulative logit link function. 
In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity. 
We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.

@@

tree.time.sp <- tree.time %>%
  group_by(Species) %>%
  dplyr::summarise(
    inc_annual_med = round(median(inc_annual, na.rm = T), 2),
    inc_annual_sd = round(sd(inc_annual, na.rm = T), 2)
  ) %>%
  arrange(inc_annual_med)
```

@@

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with `r tree.time.sp$Species[1]` having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and `r tree.time.sp$Species[nrow(tree.time.sp)]` the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.

@@

Drought sensitivity across all trees also showed negative sensitivities for both droughts with considerable variation around it (Fig 2b).

@@

@@

![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4_alternate.png)

@@

![***Figure 5: Effects of size, exposure and water availability***](display/Fig5.png)

>>>>

---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

**Title 1:** Sensitivity of tree growth to drought in seasonally dry tropical forests using long-term dendrometer band measurements

**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different droughts in a seasonally dry tropical forest

**Title 3:** Diversity of species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 4:** Long-term dendrometer band measurements reveal diverse species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests

**Title 6:**

@@

**Running title:**

**Article Type:** Letter

**Number of Words in Abstract:**  
**Number of Words in Main Text:**  
**Number of References:**  
**Number of Figures:**  
**Number of Tables:**  
**Number of Text boxes:**

**Corresponding author address:**

# Statement of Authorship {-}

# Data Accessibility Statement {-}

@@

<!-- Ecology Letters format 
-->

<!-- # Summary {-} -->
<!-- 
New Phytologist Summary format
- Rationale  
- Methods  
- Key results   
- Main conclusion   -->
# Abstract (150 words) {-}

@@

<!-- # Keywords {-} -->
\newpage

@@

<!-- Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to climate anomalies is poorly understood.  
- Woody growth has long-lasting effects on the terrestrial carbon pool but is sensitive to interannual variation in temperature and precipitation.  
- Extreme events like drought, with temperature and precipitation outside normal ranges that can lead to growth reductions that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].  
- the extent of this variation in drought responses is poorly understood.  
- with changing climate patterns across the global tropics, including novel climatic regimes [@dahinden_future_2017], there is a need to understand the drivers of this variation, to be able to predict and respond to forest responses to drought.

@@

-High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
-Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity   -->

@@

Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages between these processes. 
We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables. 
However, we also ran simpler models with species random effects only on the intercept, the results of which are reported in the Supplementary Material. 
Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as an monotonic predictor (ordered factor) and centred and scaled the other predictors. 
We modelled CII as an ordinal categorical variable with a cumulative logit link function. 
In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
<!-- Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity. 
We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.   -->

@@

tree.time.sp <- tree.time %>%
  group_by(spfull) %>%
  dplyr::summarise(
    inc_annual_med = round(median(inc_annual, na.rm = T), 2),
    inc_annual_sd = round(sd(inc_annual, na.rm = T), 2)
  ) %>%
  arrange(inc_annual_med)

sens.sp <- tree.time %>%
  filter(yr %in% c(2010, 2015)) %>%
  group_by(spfull, yr) %>%
  dplyr::summarise(
    sens_med = round(median(sens.prop, na.rm = T), 2),
    sens_mean = round(mean(sens.prop, na.rm = T), 2),
    sens_sd = round(sd(sens.prop, na.rm = T), 2)
  ) %>%
  arrange(sens_med)

# spread this for each year

sens.sp.wide <- sens.sp %>%
  ungroup() %>%
  dplyr::select(spfull, yr, sens_med) %>%
  pivot_wider(names_from = yr, values_from = sens_med, names_prefix = "sens.")

# paired correlation of 2010 and 2015 sensitivities across species

sens.sp.cor <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2015)
```

@@

The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.

@@

Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig 2b). 
Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`. 
In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively. 
In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively.

@@

```{r model summaries, echo=F, warning=F, message=F}
# load data--------------------------
datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")

tree.time <- datasets$tree.time
tree_vars <- datasets$tree_vars
sp_vars <- datasets$sp_vars

colnames(sp_vars)[1] <- "Species"

coefs_df <- readRDS("../results/models/non_negative/sensitivity_model_intercept.RData")
ranef_df <- coefs_df %>% filter(grepl("r_Species", param))
ranef_df$Species <- gsub("r_Species\\[|\\,Intercept\\]", "", ranef_df$param)

# add intercept to ranef_df
# make intercept vector
intercepts <- coefs_df %>%
  filter(param %in% "Intercept") %>%
  dplyr::select(median) %>%
  pull(median)
intercepts <- rep(intercepts, each = nrow(ranef_df) / 2)
ranef_df <- ranef_df %>%
  dplyr::mutate(
    intercept = intercepts + median,
    lwr = intercepts + lwr,
    upr = intercepts + upr
  )

# plot intercepts against species characteristics

# join ranef_df with sp_vars
ranef_df <- merge(ranef_df, sp_vars, by = "Species", all.x = TRUE)

# lms
coefs_dec_lms <- ranef_df %>%
  # filter(yr %in% yrs) %>%
  filter(Species != "ALPHVE") %>%
  nest_by(yr) %>%
  # dplyr::mutate(mod = list(lm(intercept ~ williams_dec, data = data))) %>%
  dplyr::mutate(mod = list(cor.test(data$williams_dec, data$intercept))) %>%
  dplyr::reframe(broom::tidy(mod))

isocline_coefs <- readRDS("../results/models/non_negative/coefs_isoclines_nore.RDS")
# isocline_coefs <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")

isocline_coefs <- do.call(rbind, isocline_coefs)

isocline_coefs %>% filter(param == "b_williams_dec")

# head(isocline_coefs)
```

@@

Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`) but on average had lower mean and median sensitivities in 2015 than in 2010, in alignment with median sensitivities across all trees. 
In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`). 
On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).

Across all trees, predicted sensitivities in 2010 and 2015 was heterogenous with deciduousness and TWI (Fig 3b).

Accounting for species differences in mean sensitivity, the effect of TWI on sensitivity was different in 2010 and 2015, with 2015 effects associated with deciduousness.

![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4.png)

@@

Crown exposure was associated with stronger negative sensitivities in both droughts.

Accounting for effects of exposure, remaining effects of size on sensitivity was different between the two droughts.

![***Figure 5: Effects of size, exposure and water availability***](display/Fig5_alternate2.png)
